---
title: "ETC2420 Assignment 2"
author: "Rounak Agarwal, Jonas Tiong and Daniel Klinger"
output:
  bookdown::pdf_book: null
  fig_width: 8
  toc: yes
  theme: readable
  number_sections: no
  fig_height: 5
  pdf_document: default
---

```{r setup, ,echo = FALSE, include = FALSE, warning=FALSE,error=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE,error=FALSE, message = FALSE)
options(digits = 2)
```


```{r,echo = FALSE, include = FALSE}

library(rstatix)
library(tidyverse)
library(dplyr)
library(broom)
library(tidymodels)
library(splines)
library(knitr)
library(gtsummary)
library(kableExtra)
```


```{r transforming data,echo = FALSE, include = FALSE, warning=FALSE,error=FALSE}
dat <- read_csv("https://raw.githubusercontent.com/itsrounak/Estimation-of-Neonatal-Mortality-Rate-NMR-/main/neonatal_mortality.csv")

dat <- dat %>% 
  mutate(scaled_nmr = log(nmr / (u5mr - nmr)), #scaled_nmr is the scaled NMR
         log_u5mr = log(u5mr),
         years = year - min(year)) %>% #starting from year 1951
  select(-year) %>% 
  rename(year = years) 
```
\newpage

# Question 1

# Part A{.unnumbered}
## Estimating Neonatal Mortality
### Introduction

The aim of this paper is to estimate the average neonatal mortality rate (NMR)
by a linear regression analysis. This is of interest as we want to analyse
the claim that the *composition* of child mortality has changed, where the under 
5 mortality rate (U5MR) has declined globally over the decades (since 1950, when 
our data starts) leading to a higher classification of child mortality as NMR.

Hence, our regression model(s) will look at the effect of NMR on U5MR, as well 
as other predictors such as by year and by region, if applicable.
A rudimentary check of the applicability of such a model including `year`, 
`region`, `u5mr`, requires us to start by analysing the simple linear model of 
`scaled_nmr` on `year` and adding more covariates if it improves the fit of our model, as 
determined by the adjusted $R^{2}$. That is, does adding more covariates
describe more of our child mortality data?

```{r finding our model, warning=FALSE,error=FALSE}
fit1 <- lm(scaled_nmr ~ year, data = dat)
fit2 <- lm(scaled_nmr~ year + region, data = dat)
fit3 <- lm(scaled_nmr ~ year + log_u5mr + region, data = dat)
fit4 <- lm(scaled_nmr ~ year + log_u5mr*region, data = dat)
```

```{r, warning=FALSE,error=FALSE}
model_selec <- tibble("Model" = c(1, 2,3,4), 
                      "Adjusted R Squared" = c(summary(fit1)$adj.r.squared,
                                               summary(fit2)$adj.r.squared,
                                               summary(fit3)$adj.r.squared,
                                               summary(fit4)$adj.r.squared))
model_selec%>% 
  kable(position = "center")
```
From the table, it can be seen that our first model is unable to explain most
of our data as adduced by its measly ajdusted $R^{2} = 0.04$. A visible
improvement is made to our model when we include a categorical variable that 
accounts for region ($R^{2} = 0.43$). This is further improved when we add 
the log of U5MR to our model (we must take the log of `u5mr`, as we
are, in fact, modelling a scaled NMR that is more appropriate for regression
lines - scaled NMR = $log(\frac{nmr}{u5mr - nmr})$). Finally, we consider the 
benefits of using an interaction effect between the covariates `log_u5mr` and 
`region`  through an analysis of variance that compares a model with the interaction 
to one without.

```{r, warning=FALSE,error=FALSE}
anova(fit3,fit4) 
```
According to the analysis of variance, there is reasonable evidence to conclude
at the 0.001 level of significance, that an interaction effect will improve
the fit our regression. The residual sum of squares (RSS) also has improved with
the interaction effect, which is linked to an improved fit to our data (by
the RSE).

Looking at the graphs below, we may observe that every region has its own
clear trend, a declining trend, of U5MR. And, this is at paces and level of U5MR
at each point of time quite different to every other region. This suggests that
it is appropriate to include a region variable and also model the interaction 
of region with U5MR. The second set of graphs plotting NMR v U5MR describes
a downward trend that follows in NMR with the decline in U5MR. Hence, it also 
gives reason to think we should include U5MR in our regression. 

```{r, warning=FALSE,error=FALSE}

dat %>% ggplot(aes(x = year, y = log_u5mr)) + 
  geom_point() + 
  geom_smooth(method="lm", se = FALSE) + 
  facet_wrap(~region)+ 
  labs(x = "Year ",
       y = "Scaled NMR",
       title = "Year Vs log of u5mr in different regions")+theme_bw()

dat %>% ggplot(aes(x = year, y = log_u5mr, 
                   colour = region)) + 
  geom_point() + 
  geom_smooth(method="lm", se = FALSE)+ 
  labs(x = "Year ",
       y = "Scaled NMR",
       title = "Year Vs log of u5mr")+theme_bw()

dat %>% ggplot(aes(x = log_u5mr, y = scaled_nmr)) + 
  geom_point() + 
  geom_smooth(method="lm", se = FALSE) + 
  facet_wrap(~region)+ 
  labs(x = "log(u5mr) ",
       y = "Scaled NMR",
       title = "Scaled NMR Vs log of u5mr in different regions")+theme_bw()

dat %>% ggplot(aes(x = log_u5mr, y = scaled_nmr, 
                   colour = region)) + 
  geom_point() + 
  geom_smooth(method="lm", se = FALSE)+ 
  labs(x = "log(u5mr) ",
       y = "Scaled NMR",
       title = "Scaled NMR Vs log of u5mr")+theme_bw()

```
We conclude that our chosen linear regression model is `scaled_nmr` on 
`year`, the log of `u5mr`, and `region`, where we will model, in addition to 
the effects of condition on `u5mr` and `region` on `scaled_nmr`, a separate slope
for `u5mr` for every `region` there is in the data (the interaction effect).


## Adding the model to our workflow
```{r our chosen model, warning=FALSE,error=FALSE}
dat_rec <- recipe(scaled_nmr ~ year + u5mr + region, data = dat) %>% 
  step_log(u5mr) %>% 
  step_dummy(all_nominal_predictors()) %>% 
  step_interact(terms = ~ u5mr*starts_with("region")) 
#Step_interactallows for interaction between the covariates. 
#Needed to convert all non-numeric covariates to dummy variables

# Model
lm_spec <- linear_reg() %>% 
  set_engine("lm")

# Workflow linking the recipe & model
dat_wf <- workflow() %>% 
  add_recipe(dat_rec) %>% 
  add_model(lm_spec)

# Our regression line
fit_chosen <- dat_wf %>% fit(data = dat) %>% extract_fit_parsnip()
dat_chosen <- fit_chosen$fit %>% augment()
dat_chosen <- cbind(dat %>% select(-year, -u5mr), dat_chosen)

```

## Assessing fit of all data
```{r, warning=FALSE,error=FALSE }
dat_chosen %>% ggplot(aes(x = .fitted, y = .resid)) +
  geom_point(color = "#757bc8") +
  geom_hline(yintercept = 0,
             linetype = "dashed",
             colour = "#ff8400") + 
  labs(x = "Fitted values",
       y = "Residuals",
       title = "Residuals Vs Fitted")+theme_bw()


dat_chosen %>% ggplot(aes(sample = .resid / .sigma)) + 
  geom_qq() + 
  geom_abline(intercept = 0, slope = 1) + 
  labs(x = "Theoretical quantiles", 
       y = "Standardised residuals",
       title = "Normal Q-Q Plot")

p <- 14
n <- length(dat$nmr)
threshold <- (p + 1) / n

dat_chosen %>% 
  mutate(high_cooks_d = .cooksd > threshold) %>% 
  ggplot(aes(x = .hat, y = .std.resid, 
             colour = high_cooks_d)) +
  geom_point(alpha = 0.5) +
  geom_line(aes(x = .cooksd, y = .fitted), col = "red", lty = "dashed") + 
  labs(x = "Leverage (Cook's D)",
       y = "Standardised residuals",
       title = "Residuals v Leverage (showing high leverage points)")
```
We take the residuals vs fitted; normal q-q plot; and residuals vs leverage 
as diagnostics for the fit of our data.

**Residuals vs Fitted**
A linear model is appropriate when there is a constant variance against fitted
values (of our regression). From our graph, there is a large change in variance 
after -0.5, where there are some large outliers compared to the previously 
constant variance data point before. After 0.5, the residuals sit well below 
the zero line as well. Hence, this is a signs of heretoskedasticity, 
meaning our regression line is suffering from the effects of high leverage
or outlier data points and that our plot is not perfect. It does not, however,
seem like that there is clear pattern emerging from this plot, thus 
it may stand that the residuals are independent which would be a good sign for 
this regression.

**Normal Q-Q Plot**
A linear model may still be useful for prediction even if the assumptions that
residuals are i.i.d normal do not hold (these are the assumptions 
that residuals should have constant variances and by independent in the residuals
vs fitted plot). What has to hold is that they look approximately normal, as per
the q-q plot. Hence, our regression model suffers in that the standardised 
residuals stray at the ends and beginning of the q-q plot, accounting 
for the large uncertainty (or noise) in our data in modelling the `scaled_nmr`.

**Residuals vs Leverage**
This plot is useful for identifying those data points that greatly influence 
our regression. If there is high residual (poorly described by the model) and 
high leverage for a point (high influence on the fit), then such a point is 
considered an outlier. Outliers mean our regression will suffer in fitting 
our data well.
Looking at the blue points, there are many outliers in our data, hence our
regression model is not fitting it very well.

Overall, the fit of model is not perfect, however, its applicability for 
prediction is not solely determined by these diagnostic plots. Moreover, 
despite the inadequacies, the fit is not terrible as there is no clear pattern 
in the fitted vs residuals plot for instance (suggesting we are using the 
wrong type of model).


## Fitting model by region
```{r , warning=FALSE,error=FALSE}
dat_chosen %>% ggplot(aes(x = .fitted, y = .resid)) + 
  geom_point(color = "#757bc8") +
  geom_hline(yintercept = 0,
             linetype = "dashed",
             colour = "#ff8400") + 
  facet_wrap(~region) + 
  labs(x = "Fitted values",
       y = "Residuals",
       title = "Residuals v Fitted")+theme_bw()

dat_chosen %>% ggplot(aes(sample = .resid / .sigma)) + 
  geom_qq() + 
  geom_abline(intercept = 0, slope = 1) + 
  facet_wrap(~region) +
  labs(x = "Theoretical quantiles", 
       y = "Standardised residuals",
       title = "Normal Q-Q Plot")+theme_bw()

p <- 14
n <- length(dat$nmr)
threshold <- (p + 1) / n

dat_chosen %>% 
  mutate(high_cooks_d = .cooksd > threshold) %>% 
  ggplot(aes(x = .hat, y = .std.resid, 
             colour = high_cooks_d)) +
  geom_point(alpha = 0.5) +
  geom_line(aes(x = .cooksd, y = .fitted), col = "red", lty = "dashed") + 
  facet_wrap(~region) + 
  labs(x = "Leverage (Cook's D)",
       y = "Standardised residuals",
       title = "Residuals Vs Leverage (showing high leverage points)")+theme_bw()
```
The regression fares very well for South Asia and not so well for the other regions.
By region, the residuals seem very high at some fitted values and for most part 
are constant in variance as it was for the whole data set. Visible differences
are for the region Eurasia (Central Europe, Eastern Europe, Central Asia) and 
the high income countries. This will require further exploration, however,
we think that in the case of Eurasia, this may be caused by the demographic 
isolation and make-up of some settlements being far away from each other and 
the central city (giving rise to the poor fit). Such factors would lead 
to increased risk of death of children, and particularly for young babies who 
need intensive care that may not be able to provided in such places. In the case
of high income countries, fitted values tend to center, we believe, because they
generally have a minimum understanding of where they want there child mortality rates
to be at, and its correlated to the high expenditure and wealth of these countries 
to support such rates. Moreover, the high residuals may be caused from the fact 
some of these countries have very strong policies and institutions that prevent
neonatal deaths. What is clear is that these regions, and with the issue of 
heteroskedaticity for most of the other regions, make our fit to the data limited
by our model.
Such issues can be seen by the normalised q-q plot, where residuals stray at the 
tails and the high number of outliers (save South Asia).
It will also be interesting to explore why South Asia is fitting our regression 
model very well. This may, however, be largely due to the lack of outliers 
in the data.


## Fitting model by the three selected countries
```{r, warning=FALSE,error=FALSE }
countries <- c("Malaysia", "Brazil", "Australia")
dat_chosen_three <- dat_chosen %>% 
  filter(country_name %in% countries) #IS THERE A BETTER CHOICE OF COUNTRIES?

dat_chosen_three %>%
  ggplot(aes(x = year, y = .resid)) +
  geom_point(alpha = 0.5,color = "blue") +
  geom_hline(yintercept = 0,
             linetype = "dashed",
             colour = "#ff8400")+ 
  facet_wrap(~country_name, nrow = 3) + 
  labs(x = "Fitted values",
       y = "Residuals",
       title = "Residuals Vs Fitted")+theme_bw()


dat_chosen_three %>% ggplot(aes(sample = .resid / .sigma)) + 
  geom_qq() + 
  geom_abline(intercept = 0, slope = 1) + 
  facet_wrap(~country_name) +
  labs(x = "Theoretical quantiles", 
       y = "Standardised residuals",
       title = "Normal Q-Q Plot")+theme_bw()



dat_chosen_three %>% 
  mutate(high_cooks_d = .cooksd > threshold) %>% 
  ggplot(aes(x = .hat, y = .std.resid, 
             colour = high_cooks_d)) +
  geom_point(alpha = 0.5) +
  geom_line(aes(x = .cooksd, y = .fitted), col = "red", lty = "dashed") + 
  facet_wrap(~country_name) + 
  labs(x = "Leverage (Cook's D)",
       y = "Standardised residuals",
       title = "Residuals Vs Leverage (showing high leverage points)") +
  theme_bw()
```
We have chosen Australia, Brazil, and Malaysia. The reason being is that they 
are part of three different continents and are in very different stages of their growth. 
Australia is a high income country, Brazil and Malaysia are not.
There is a pattern for all these countries in their residuals, which means the 
regression does not do well to model these countries' NMR. This is, however, 
not the biggest of concern, for each country will likely have huge variance and 
its own distinct pattern for NMR over time compared to global averages and regional
averages. 
The residuals are clearly not normal as per the q-q plot. None of these 
countries have outliers or even high leverage points. Yet, there standardised 
residuals is quite high.


## Validation Set Approach to RMSE & MAE

We are concerned here to estimate the test root mean square error (RMSE) and 
test mean absolute error (MAE), through the validation set approach.
A low estimate for these metrics would be a favourable sign to the fit of our 
model.

```{r, warning=FALSE,error=FALSE }
dat_split <- initial_split(data = dat, strata = region)
dat_train <- training(dat_split)
dat_valid <- testing(dat_split)

fit_chosen_train <- dat_wf %>% fit(dat_train)
dat_chosen_valid <- dat_valid %>% 
  bind_cols(predict(fit_chosen_train, dat_valid))

dat_chosen_valid %>% metrics(truth = scaled_nmr, estimate = .pred) %>% 
  rename("Estimator" = ".metric", "Estimate"= ".estimate")%>%
  select(-`.estimator`)%>%
  kable()
```
We opted for the default 75 per cent split of our data to training and 
validation sets. On this, our model (on the training set) has an estimated
test RMSE of 0.42 and an estimated test MAE of 0.30. Such values suggest 
that our model is not doing too bad in fitting our data, with its estimated 
test MAE of 0.30 being favourable sign.

## Prediction CLT intervals
We test here how well our model can predicate the data despite the inadequancies
in its fit as determined and discussed before.


## Prediction CLT interval for whole data
```{r, warning=FALSE,error=FALSE }
pred_nmr <- predict(fit_chosen$fit, dat_chosen, interval = "prediction") %>% 
  as.data.frame()

pred_nmr <- bind_cols(dat, pred_nmr) %>% mutate(
  fitted = exp(fit) * u5mr / (1 + exp(fit)), 
  upr_ci = exp(upr) * u5mr / (1 + exp(upr)), 
  lwr_ci = exp(lwr) * u5mr / (1 + exp(lwr))
  ) %>% select(-fit, -upr, -lwr)

pred_nmr_all <- pred_nmr %>% group_by(year) %>% 
  summarise(fit = mean(fitted), 
            lwr = mean(lwr_ci), 
            upr = mean(upr_ci),
            nmr = nmr)

pred_nmr_all %>% ggplot(aes(x = year)) +
  geom_line(aes(y = nmr), colour = "grey") +
  geom_line(aes(y = fit), colour = "#ff8400") +
  geom_line(aes(y = lwr), colour = "blue", linetype = "dashed") +
  geom_line(aes(y = upr), colour = "blue", linetype = "dashed") +
    labs(
      x = "Years",
      y = "Neonatal Mortality Rate (NMR)", 
      title = "Neonatal Mortality Rate per annum")+theme_bw()
```
The grey lines are the plot of the points from the data, and the yellow 
line is the average NMR as predicted by our model. The blue lines that encloses 
this is the 95 per cent prediction CLT interval.
From the graph, we notice that again the downward trend in NMR over the years.
We also see that our model predicts an average NMR with its
prediction intervals that sits within the true values (the grey lines). 
This suggests that the fit of model is good on the scale of global averages.


## Prediction CLT interval by region
```{r, warning=FALSE,error=FALSE }
pred_nmr_region <- pred_nmr %>% group_by(year, region) %>% 
  summarise(fit = mean(fitted), 
            lwr = mean(lwr_ci), 
            upr = mean(upr_ci))

pred_nmr %>% ggplot(aes(x = year)) +
  geom_line(aes(y = nmr, group = country_name), 
            colour = "grey") +
  geom_line(data = pred_nmr_region, aes(y = fit), 
            colour = "#ff8400") +
  geom_line(data = pred_nmr_region, aes(y = lwr), 
            colour = "blue", 
            linetype = "dashed") +
  geom_line(data = pred_nmr_region, aes(y = upr), 
            colour = "blue", 
            linetype = "dashed") +
  facet_wrap(~region, nrow = 3) + 
  labs(
      x = "Years",
      y = "Neonatal Mortality Rate (NMR)", 
      title = "Neonatal Mortality Rate per annum")+theme_bw()
```
By region, the breakdown isn't as favourable. Our model predicts large 
uncertainty in its value for NMR for South Asia, which comes from uncertainty 
in other parts of the whole data set. This is not bad per se, as the future trend 
of  NMR is reliant on the observations of the trajectory of NMR in the past for 
other countries.
For North Africa / the Middle East and Sub-Saharan Africa, the uncertainty is 
also quite big in the predicted values, which is a reflection of the underlying
uncertainty of estimating NMR for them.  What is surprising is that for high income
countries and Eurasia, this uncertainty is quite low. Perhaps this is 
because for our q-q plots in ยง2.2, the residuals only strayed very far off in the
tails, which meant that our model was still good for predicting the average NMR.


## Prediction CLT intervals of three countries
```{r, warning=FALSE,error=FALSE }
pred_nmr_three <- pred_nmr %>% filter(country_name %in% countries)


pred_nmr_three <- pred_nmr_three %>% group_by(year, country_name) %>% 
  summarise(fit = mean(fitted), 
            lwr = mean(lwr_ci), 
            upr = mean(upr_ci),
            nmr = nmr)

pred_nmr_three %>% ggplot(aes(x = year)) +
  geom_line(aes(y = nmr), colour = "grey") +
  geom_line(aes(y = fit), colour = "#ff8400") +
  geom_line(aes(y = lwr), colour = "blue", linetype = "dashed") +
  geom_line(aes(y = upr), colour = "blue", linetype = "dashed") + 
  facet_wrap(~country_name, nrow = 3) + 
  labs(
      x = "Years",
      y = "Neonatal Mortality Rate (NMR)", 
      title = "Neonatal Mortality Rate per annum")+theme_bw()
```
For these three countries, the prediction intervals has large uncertainty that 
is bigger than the true values. This again is expected, we cannot expect that 
our model will fare just as well as it did on the whole and for each region as it 
would for just one country.


\newpage

# Part B{.unnumbered}
## Non-linearity of U5MR vs NMR
A reason that our fit for our linear regression model was not very good could 
have possible due to not capturing the relationship between U5MR and NMR in 
the best way. We consider the linear regression against a non-linear regression
model using a B-Spline for the `log_u5mr` covariate. 

```{r}
dat  %>%
  ggplot(aes(x = log_u5mr, y = scaled_nmr)) +
  geom_point() + 
  geom_smooth(method = "lm", formula = y ~ bs(x), colour = "blue") +
  geom_smooth(method = "lm", colour = "red") +
  facet_wrap(~region, nrow = 3) + 
  labs(x = "Log-scaled U5MR", 
       y = "Scaled NMR",
       title = "Scaled Neonatal Mortality Rate (NMR) vs Log-scaled Under 5 
       Mortality Rate (U5MR)")
```

This model has visibile improvements over its linear counterparts for some regions.
Latin America and Caribbean; North Africa / Middle East; Eurasia (Central Europe, 
Eastern Europe, Central Asia); High income are regions where the B-spline 
modelling has reduced the distance of the residuals from the fitted values 
(the blue line). Other countries have had muted benefits to the linear model (the 
red line), where in South Asia there is no improvements. As noted in ยง2.2, the 
model has plotted the relationship very well for South Asia and South Asia benefits
from little outliers. Hence, it makes sense that the B-Spline has opted 
to leave it as a linear model for that region. 
We thus conclude that it will be of interest to model and compare a regression 
model using B-spline for the log of `u5mr` to estimate `scaled_nmr`.

In this second model, we consider the non linear affect that U5MR has on NMR

```{r, echo=FALSE,include=FALSE, warning=FALSE,error=FALSE}
# reloading data anew
dat <- read_csv("https://raw.githubusercontent.com/itsrounak/Estimation-of-Neonatal-Mortality-Rate-NMR-/main/neonatal_mortality.csv")
dat <- dat %>% 
  mutate(scaled_nmr = log(nmr / (u5mr - nmr)), #scaled_nmr is the scaled NMR
         log_u5mr = log(u5mr),
         years = year - min(year)) %>% #starting from year 1951
  select(-year) %>% 
  rename(year = years) 
dat_split <- initial_split(data = dat, strata = region)
dat_train <- training(dat_split)
dat_valid <- testing(dat_split)
```

## Calculating DF parameter for B-Spline Function 
```{r our chosen model with B-spline for log ofu5mr,warning=FALSE, error=FALSE}

dat_rec <- recipe(scaled_nmr ~ year + u5mr + region, data = dat_train) %>% 
  step_log(u5mr) %>% 
  step_dummy(all_nominal_predictors()) %>% 
  step_interact(terms = ~ u5mr*starts_with("region")) %>% 
  step_bs(u5mr, deg_free = tune()) # using B-spline for the (log of) u5mr

lm_spec <- linear_reg() %>% 
  set_engine("lm")

dat_wf <- workflow() %>% 
  add_recipe(dat_rec) %>% 
  add_model(lm_spec)

# grid for which we will test out by CV which flexibility (1=<DF=<30) is best
df_grid <- grid_regular(deg_free(range = c(1,30)), levels = 30)

# 10-fold CV
folds <- vfold_cv(dat, v = 10, strata = region)

# tuning our model by 10-fold CV going through the grid
fit_tuned <- dat_wf %>% 
  tune_grid(resamples = folds, 
            grid = df_grid)

# choosing the best DF by selecting the lowest RMSE
lowest_rmse <- fit_tuned %>% select_best("rmse")

# finalizing the workflow on this chosen DF
final <- finalize_workflow(dat_wf, lowest_rmse)

fit_bs <- final %>% fit(dat) %>% extract_fit_parsnip() # our regression
dat_bs <- fit_bs$fit %>% augment()
dat_bs <- cbind(dat %>% select(-year), 
                dat_bs) # binding back parts of the data that was removed
```

## Choosing the DF for our B-spline for log_u5mr covariate.

We have chosen the appropriate number of basis functions for our B-spline 
regression model through tuning our model through a grid of values (for the 
degrees of freedom it can test which has the lowest RMSE. The best one 
will minimise the RMSE and $R^{2}$.

```{r, warning=FALSE,error=FALSE}
fit_tuned %>% collect_metrics() %>%
  ggplot(aes(x = deg_free, y = mean,
             colour = .metric)) +
  geom_errorbar(aes(ymin = mean - std_err,
                    ymax = mean + std_err),
                alpha = 0.5) +
  geom_line() +
  facet_wrap(~.metric, scales = "free", nrow = 2) + 
  scale_x_log10() + 
  theme(legend.position = "none") + 
  labs(x = "Degree of freedom (DF)", 
       y = "Mean")+theme_bw()
```

## Fitting the  Models considering the non-linear effect and comparing them
```{r, warning=FALSE,error=FALSE}
fit4 <- lm(scaled_nmr ~ year + log_u5mr*region, 
           data = dat)
fit5 <- lm(scaled_nmr ~ year + bs(log_u5mr, df = lowest_rmse$deg_free)*region, 
           data = dat)
anova(fit4, fit5)
```
Doing an analysis of variance, it can be seen that there is reasonable evidence 
to conclude that there is an improvement using the B-spline for the log `u5mr`
in our model, at the 0.001 level of significance.

## New fit for all data
```{r checking the fit of our chosen model, warning=FALSE,error=FALSE}
dat_bs %>% ggplot(aes(x = .fitted, y = .resid)) +
  geom_point(color = "#757bc8") +
  geom_hline(yintercept = 0,
             linetype = "dashed",
             colour = "#ff8400") + 
  labs(x = "Fitted values",
       y = "Residuals",
       title = "Residuals v Fitted")+theme_bw()

dat_bs %>% ggplot(aes(sample = .resid / .sigma)) + 
  geom_qq() + 
  geom_abline(intercept = 0, slope = 1) + 
  labs(x = "Theoretical quantiles", 
       y = "Standardised residuals",
       title = "Normal Q-Q Plot")+theme_bw()

p <- 14
n <- length(dat$nmr)
threshold <- (p + 1) / n 

dat_bs %>% 
  mutate(high_cooks_d = .cooksd > threshold) %>% 
  ggplot(aes(x = .hat, y = .std.resid, 
             colour = high_cooks_d)) +
  geom_point(alpha = 0.5) +
  geom_line(aes(x = .cooksd, y = .fitted), col = "red", lty = "dashed") + 
  labs(x = "Leverage (Cook's D)",
       y = "Standardised residuals",
       title = "Residuals v Leverage (showing high leverage points)")+theme_bw()
```

In terms of the fitted vs residual plot, there is prima facie not many signs that 
there has been a significant improvement. However, near the 0.5 value, the residuals 
are  now more spread out in each side of the zero line, something that was noted 
to be lacking in the linear model. This is a favourable sign of improvement 
in the fit of our model.  Moreover, looking carefully, more points lie in both 
sides of the zero line in equal amounts for the entirely of this plot than before. 
Hence, this suggests that the variances is more constant and independent for this 
model that the old one. 
The normal q-q plot, however, does appear to be the same. So, this suggests that 
the estimating accuracy of this model is not that different to the old linear 
model.
There is also less outlier points under this model. Hence, this model is not 
severely effected as before by different data points, improving the fit to the 
data.


## New fit for all data by region
```{r checking the fit by region, warning=FALSE,error=FALSE}
dat_bs %>% ggplot(aes(x = .fitted, y = .resid)) + 
  geom_point(color = "#757bc8") + 
  geom_hline(yintercept = 0, 
             linetype = "dashed", 
             colour = "#ff8400") + 
  facet_wrap(~region) + 
  labs(x = "Fitted values",
       y = "Residuals",
       title = "Residuals v Fitted")+theme_bw()

dat_bs %>% ggplot(aes(sample = .resid / .sigma)) + 
  geom_qq() + 
  geom_abline(intercept = 0, slope = 1) + 
  facet_wrap(~region) + 
  labs(x = "Theoretical quantiles", 
       y = "Standardised residuals",
       title = "Normal Q-Q Plot")+theme_bw()

dat_bs %>% 
  mutate(high_cooks_d = .cooksd > threshold) %>% 
  ggplot(aes(x = .hat, y = .std.resid, 
             colour = high_cooks_d)) +
  geom_point(alpha = 0.5) +
  geom_line(aes(x = .cooksd, y = .fitted), col = "red", lty = "dashed") + 
  facet_wrap(~region) +
  labs(x = "Leverage (Cook's D)",
       y = "Standardised residuals",
       title = "Residuals v Leverage (showing high leverage points)")+theme_bw()
```

By region, we can see again that the fit has improved. For high income countries
and Euarasia, the data is very more spread out across the zero line than it was 
under the linear model. Again, the q-q plot suggests that the predicting 
accuracy will not be significantly different. 
For the number of outliers, this has reduced for some regions, but across the 
board, the diagnostic plot shows that it has stayed relatively the same. 
All in all, these plots suggests that there is indeed benefit to using a non-linear 
model for estiamting the average NMR.


## New fit for the data by the selected countries
```{r checking the fit by three countries, warning=FALSE,error=FALSE}
dat_bs_three <- dat_bs %>% 
  filter(country_name == "Australia" | 
           country_name == "Brazil" | 
           country_name == "France")

dat_bs_three %>%
  ggplot(aes(x = year, y = .resid)) +
  geom_point(alpha = 0.5, color = "blue") +
  geom_hline(yintercept = 0,
             linetype = "dashed",
             colour = "#Ff8400") + 
  facet_wrap(~country_name, nrow = 3) + 
  labs(x = "Fitted values",
       y = "Residuals",
       title = "Residuals v Fitted")+theme_bw()
```

Plotting for our chosen three countries, there is still a pattern for the 
residuals vs fitted, however, the data is now more spread across the zero line. 
As discussed before, this pattern is expected for each country will be different
to its regional and the global average in child mortality data. So, we 
sacrifice the ability to plot any abitrary country for the ability to predict 
on the regional and global scale NMR. Likewise, this note rings true to the 
q-q plot which shows that at the tails the data points differing the theoretical 
normal quantiles (suggesting a lack of fit).
A concern for using B-spline is that, however contrary to the data, the predicted 
leverage (as per Cook's D) is vastly different for these countries. However, 
such concerns are slightly alleviated by the fit by region, for the plot fares 
well in those. 



## Validation set approach to estimating the test RMSE & MAE
```{r validation set approach to estimating the test RMSE and MAE, warning=FALSE,error=FALSE}
fit_bs_train <- final %>% fit(dat_train)
dat_bs_valid <- dat_valid %>% 
  bind_cols(predict(fit_bs_train, dat_valid))

dat_bs_valid %>% metrics(truth = scaled_nmr, estimate = .pred) %>%
  select(-`.estimator`)%>%
  kbl(booktabs = T) %>%
  kable_styling(position = "center")
```

Using the validation set approach, splitting our data again in 75-25 per cent 
(by region) as per ยง2.4, and running the model (fitted to the training set) on 
the validation set, we see that the estimated test RMSE and MAE is approximately 
the same as under the linear model. This resulted is not suprising for the q-q 
plots aforementioned suggested that for predicting, our model will not be 
significally different. 

## Prediction CLT intervals
We test here how well our model can predict the data compared to the old 
model.

## Prediction CLT intervals for whole data
```{r prediction CLT interval for whole data, warning=FALSE,error=FALSE}
pred_nmr <- predict(fit_bs$fit, dat_bs, interval = "prediction") %>% 
  as.data.frame()

pred_nmr <- bind_cols(dat, pred_nmr) %>% mutate(
  fitted = exp(fit) * u5mr / (1 + exp(fit)), 
  upr_ci = exp(upr) * u5mr / (1 + exp(upr)), 
  lwr_ci = exp(lwr) * u5mr / (1 + exp(lwr))
  ) %>% select(-fit, -upr, -lwr)

pred_nmr_all <- pred_nmr %>% group_by(year) %>% 
  summarise(fit = mean(fitted), 
            lwr = mean(lwr_ci), 
            upr = mean(upr_ci),
            nmr = nmr)

pred_nmr_all %>% ggplot(aes(x = year)) +
  geom_line(aes(y = nmr), colour = "grey") +
  geom_line(aes(y = fit), colour = "red") +
  geom_line(aes(y = lwr), colour = "blue", linetype = "dashed") +
  geom_line(aes(y = upr), colour = "blue", linetype = "dashed") +
    labs(
      x = "Years",
      y = "Neonatal Mortality Rate (NMR)", 
      title = "Neonatal Mortality Rate per annum")+theme_bw()
```

According to the plot below, our predicted values and the 95 per cent prediction 
CLT interval is very similar to the one produced by the linear model. 
This is again suggested by the q-q plots before. Hence, for estimating our data, 
if by global average for NMR, then either models will do.


## Prediction CLT intervals by region
```{r prediction CLT by region, warning=FALSE,error=FALSE}
pred_nmr_region <- pred_nmr %>% group_by(year, region) %>% 
  summarise(fit = mean(fitted), 
            lwr = mean(lwr_ci), 
            upr = mean(upr_ci))

pred_nmr %>% ggplot(aes(x = year)) +
  geom_line(aes(y = nmr, group = country_name), 
            colour = "grey") +
  geom_line(data = pred_nmr_region, aes(y = fit), 
            colour = "red") +
  geom_line(data = pred_nmr_region, aes(y = lwr), 
            colour = "blue", 
            linetype = "dashed") +
  geom_line(data = pred_nmr_region, aes(y = upr), 
            colour = "blue", 
            linetype = "dashed") +
  facet_wrap(~region, nrow = 3) + 
  labs(
      x = "Years",
      y = "Neonatal Mortality Rate (NMR)", 
      title = "Neonatal Mortality Rate per annum")+theme_bw()
```

By region, this stays the same, that the prediction value and the uncertainty 
in that (by the blue prediction interval) is very similar to the linear model one.


## Prediction CLT intervals by three countries
```{r prediction CLT by three countries, warning=FALSE,error=FALSE}
pred_nmr_three <- pred_nmr %>% filter(country_name %in% countries)


pred_nmr_three <- pred_nmr_three %>% group_by(year, country_name) %>% 
  summarise(fit = mean(fitted), 
            lwr = mean(lwr_ci), 
            upr = mean(upr_ci),
            nmr = nmr)

pred_nmr_three %>% ggplot(aes(x = year)) +
  geom_line(aes(y = nmr), colour = "grey") +
  geom_line(aes(y = fit), colour = "red") +
  geom_line(aes(y = lwr), colour = "blue", linetype = "dashed") +
  geom_line(aes(y = upr), colour = "blue", linetype = "dashed") + 
  facet_wrap(~country_name, nrow = 3) + 
  labs(
      x = "Years",
      y = "Neonatal Mortality Rate (NMR)", 
      title = "Neonatal Mortality Rate per annum")+theme_bw()
```

Even by the three countries, this looks the same still. Hence, we conclude 
that for estimating, either model will do, though the B-spline model 
leads to lesser residuals for some regions and thus is still favourable in that 
regard.

# Summary of Findings & Conclusion {.unnumbered}

We have modelled the covariates of `region`, the log of `u5mr`, and `year` to 
the `scaled_nmr`, with an interaction effect for `u5mr` at every `region`. Such 
a model we found to be the best. It was also found that, in terms of outputted 
residuals, that having a B-Spline for `u5mr` was favourable for there was some 
non-linearity in this variable against `u5mr`. However, overall, the linear 
model and the non-linear one was very similar to each other in predicting 
the average NMR. We also saw that there was a downward trend for all countries 
in their NMR, and this was evidenced to be due to the changing composition of 
mortality that we wanted to test. Further exploration of some of this is, however, 
needed for each region, as some regions were different in their pace and value 
of NMR to others.
Our models suffered in the lack of fit across differing regions, notably those 
in high income countries and in Eurasia.
In conclusion, for estimating the average NMR, although the B-Spline model showed 
clear improvement in its analysis of a variance and evidenced in its fitted 
vs residual plot, there is little other evidence to suggest that we should 
favour it for linear model in prediction.